Innovation and elaboration on the avian tree of life - supplementary materials

### 2022-10-18 Manuscript/main_flipped_narrative.tex
Update main_flipped_narrative.tex

% paragraph 1:  introducing the major axis of variation + novelty and innovation
Evolutionary theory predicts that over short (microevolutionary) timescales phenotypic evolution will tend to follow evolutionary lines of genetic least resistance \cite{schluter1996adaptive}. Over longer timescales, the evolutionary line of least resistance is analogous to the major axis of phenotypic variation \cite{marroig2005size,fasanelli2022allometry}. Numerous studies show that phenotypic divergence is biased along the genetic line of least resistance, and therefore reflects the major axis of phenotypic variation, over 1-2 million year timescales. Recent evidence even suggests that this stability can extend over macroevolutionary timescales spanning tens of millions of years \cite{mcglothlin2018adaptive}. However, while the microevolutionary line of least resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic divergence is an emergent property of genomic and developmental constraints interacting with selection on a moving the adaptive landscape \cite{jones2004evolution}. The widespread evidence of major shifts in phenotypes across the tree of life \cite{pagel2022general,cooney2017mega,venditti2011multiple,khabbazian2016fast,smaers2021evolution} suggests a tension in the directions of phenotypic evolution across scales where macroevolutionary divergence along lines of least resistance gives way to mega-evolutionary divergence leading to distinct phenotypes or directions of evolution. Resolving this tension requires an understanding of the balance between deep-time reorientation of phenotypic major axes of evolution among clades and the directions of evolution of species within clades.

% paragraph 2:  introducing the Endler view.
 Endler et al. \cite{endler2005animal}, writing in the specific context of signal evolution, describe two  routes to diversity that we suggest can help to illuminate this challenge. Specifically, Endler et al. refer to evolution of new phenotypes along the same axis as \textit{elaboration}. This is akin to phenotypic divergence among species following lines of least resistance, and, by definition, would be expected to be a major driver of phenotypic evolution. In contrast, they refer to evolution away from the major axis as \textit{innovation}. More specifically they describe new phenotypes falling "off the line in random directions" as \textit{random innovation} and new phenotypes falling "off the line in only one direction (or a few)" as \textit{specific innovation}.  For example, studies of allometry in which the allometric slope varies among clades \cite{marroig2005size,Rombaut2022} are consistent with specific innovation. Relative to elaboration, both random and specific innovation may be more likely to generate phenotypic novelty, but this need not be synonymous with being a primary driver of phenotypic diversity at broad scales. These core concepts provide a powerful theoretical framework with which to study macro to mega-evolutionary divergence and in particular to determine the relative importance of: i) idiosyncratic species-level change (random innovation) against clade-level reorientation of the directions of divergence (specific innovation), and ii) evolution along versus away from major axes of divergence. The contributions of these multivariate evolutionary pathways are poorly-known despite their importance for understanding the origins of diversity from macro- to mega-evolutionary scales. 
 Here we assess the stability of trait correlations across the global radiation of birds and estimate and compare major axes of beak shape evolution across lineages and scales to address the contributions of elaboration and innovation to the generation of avian phenotypic diversity.

% paragraph 3: how we measure this (glmm) on the birds data + nestedness
\subsection{Modelling nested covariances}
Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires
1) large, multivariate datasets to allow exploration of trait covariances at different scales,
2) reliable and efficient computational methods to estimate the major axes of phenotypic divergence,
and 3) a set of mathematical tools that can estimate degrees of elaboration and both random and specific innovation at any scale. To meet the first challenges we use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}). To meet the second two challenges we propose a novel analytical pipeline for measuring elaboration and random innovation at the species level and both elaboration and specific innovation at the clade level (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}). In order to provide a baseline on which to calculate elaboration and innovation we use phylogenetic generalised mixed effect models (pGLMM) with beak shape as an eight-dimensional response variable (see Methods; \cite{MCMCglmm}). Such models allow us to measure variance-covariance matrices partitioned into phylogenetic and non-phylogenetic components. From these models we can calculate major axes of divergence as the leading eigenvector in trait space. We regard the major axis of the phylogenetic partition as a deep-time analogue of the microevolutionary \textbf{G} matrix and posit that it represents the line of least evolutionary resistance capturing the effects of historical contingency on multivariate evolution. The major axis can be defined at any scale (i.e. for all species in the phylogeny, or for all species within any consistent clade). In order to assess the roles of elaboration and innovation across scales we fit our pGLMM as nested models in which we define variance-covariance matrices (and therefore major axes of divergence) for i) the phylogenetic partition of all species, ii) all species within each of nine clades mapping approximately to superorders, and iii) all species within each of XX clades mapping to taxonomic orders. These nested partitions of multivariate trait space provide the basis for all subsequent analyses.

\subsection{Elaboration dominates species-level divergence}
% paragraph 4: results for elaboration and innovation for species
We then measure innovation and elaboration as the multilinear algebra projection of each species onto their nested major axis of variation.
Here are the results for each species. \ref{Fig:phylogeny}
There seems to be a nested structure of the variation in novelty %TODO: needs to introduce nestedness; %TODO: needs to introduce novelty.
There is more variation the deeper you baseline comparison in the tree is, suggesting that novelty is relative to the baseline comparison.
Also, the majority of that novelty is through elaboration.
We found that high levels of elaboration are more common across all birds than high levels of innovation (Fig. \ref{Fig:phylogeny}).
However, we also found various combinations of elaboration and innovation across birds.
For example, compared to the phylogenetic major axis, the Trochilidae (hummingbirds, with long narrow beaks) display high levels of elaboration and low innovation whereas the Bucerotidae (hornbills, with large and oddly shaped beaks) display high levels of innovation and low elaboration.
Other clades such as the Apodidae (swifts, with large but gaping beaks) and the distantly related but convergent Hirundidae (swallows and martins) display high levels of both elaboration and innovation.
Overall, we find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp?
The variability of, and apparent disconnect between, innovation and elaboration holds whether comparisons are made at the order, superorder, or class-wide phylogenetic level (Fig. \ref{Fig:phylogeny})%{#fig_correlations}),%TODO: maybe add to supp?
 and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
The dominance of elaboration when measured at the species level is consistent with ideas of evolution along lines of least evolutionary resistance, yet contrasts with our observations of frequent innovation via reorientation of trait space within clades.
\textbf{But there is a paradox here:}
If most organisms evolve incrementally along major axis of variation then how can we explain the staggering diversity of species morphologies across the tree of life?

@@

% Paragraph 5+: Some elluipses stuff with bird names.
However, this change in the line of least resistance's orientation between different clade levels varies between clades (Fig.\ref{Fig:ellipses}).
While orthogonality provides a simple description of the degree of reorientation of phenotypic axes (measuring the two-dimensional plane of an angle), it does not fully capture the multi-dimensionality of the data.
To further quantify the tension implied between theory on lines of least evolutionary resistance and observations of deep-time reorientation of trait covariance, we measured the elaboration and innovation of each different clades (\textit{cf.} the elaboration and innovation of a species - see below).
As noted above, elaboration describes multivariate evolution that is primarily along the phylogenetically constrained phenotypic major axis whereas innovation describes multivariate evolution that is primarily away from the phylogenetically constrained phenotypic major axis \cite{endler2005animal}.
Clades (or species) can be both elaborators and innovators.
To measure innovation and elaboration we translated the major axes of each clade's (i.e. superorders and orders) phylogenetic variance-covariance matrices onto the global phylogenetic major axis so that they shared the same origin in the shapespace.
We then used multilinear algebra to project a focal clade's major axis onto the phylogenetic major axis (Fig.S\ref{Fig:cheat_sheet}).
We interpret the measured projection (distance onto the phylogenetic major axis) as the clade's elaboration score and the measured rejection (distance from the major axis) as the clade's innovation score.
These scores were measured for each of 4000 posterior pairs of clade vs. phylogenetic variance-covariance matrix (Fig.\ref{Fig:ellipses} boxplots).
We found that for most clades the phenotypic major axis is not aligned with the global phylogenetic major axis, (Fig. \ref{Fig:ellipses}).
Remarkably, more than half of the assessed clades (4/8 superorders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to theory \cite{schluter1996adaptive,marroig2005size}, innovation is a more common generator of morphological diversity than elaboration.
However, there is substantial variation in terms of elaboration and innovation across clades including relatively low elaboration and innovation (e.g. Cuculiformes; Fig. \ref{Fig:ellipses}), low elaboration and high innovation (e.g. Galliformes; Fig. \ref{Fig:ellipses}), high elaboration and low innovation (e.g. Coraciiformes; Fig. \ref{Fig:ellipses}) or both relatively high elaboration and innovation (e.g. Podicipediformes; Fig. \ref{Fig:ellipses}).
These patterns hold across scales including within superorders and at finer scales within the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes}).
This suggests that divergence away from line of least evolutionary resistance (i.e. the phylogenetic major axis) is a common feature of avian beak evolution across scales and further supports the suggestion of a remarkable flexibility in the direction of evolution in deep-time even in a developmentally constrained trait.

>>>>

% paragraph 1:  introducing the major axis of variation + novelty and innovation
Evolutionary theory predicts that over short (microevolutionary) timescales phenotypic evolution will tend to follow evolutionary lines of genetic least resistance \cite{schluter1996adaptive}.
Over longer timescales, the evolutionary line of least resistance is analogous to the major axis of phenotypic variation \cite{marroig2005size,fasanelli2022allometry}.
Numerous studies show that phenotypic divergence is biased along the genetic line of least resistance, and therefore reflects the major axis of phenotypic variation, over 1-2 million year timescales.
Recent evidence even suggests that this stability can extend over macroevolutionary timescales spanning tens of millions of years \cite{mcglothlin2018adaptive}.
However, while the microevolutionary line of least resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving the adaptive landscape \cite{jones2004evolution}.
The widespread evidence of major shifts in phenotypes across the tree of life \cite{pagel2022general,cooney2017mega,venditti2011multiple,khabbazian2016fast,smaers2021evolution} suggests a tension in the directions of phenotypic evolution across scales where macroevolutionary divergence along lines of least resistance gives way to mega-evolutionary divergence leading to distinct phenotypes or directions of evolution.
Resolving this tension requires an understanding of the balance between deep-time reorientation of phenotypic major axes of evolution among clades and the directions of evolution of species within clades.

% paragraph 2:  introducing the Endler view.
Endler et al. \cite{endler2005animal}, writing in the specific context of signal evolution, describe two routes to phenotypic novelty that we suggest can help to illuminate this challenge.
Specifically, Endler et al. refer to evolution of new phenotypes along the same axis as \textit{elaboration}.
This is akin to phenotypic divergence among species following lines of least resistance, and, by definition, would be expected to be a major driver of phenotypic evolution.
In contrast, they refer to evolution away from the major axis as \textit{innovation}.
More specifically they describe new phenotypes falling "off the line in random directions" as \textit{random innovation} and new phenotypes falling "off the line in only one direction (or a few)" as \textit{specific innovation}.
For example, studies of allometry in which the allometric slope varies among clades \cite{marroig2005size,Rombaut2022} are consistent with specific innovation.
Relative to elaboration, both random and specific innovation may be more likely to generate phenotypic novelty, but this need not be synonymous with being a primary driver of phenotypic diversity at broad scales.
These core concepts provide a powerful theoretical framework with which to study macro to mega-evolutionary divergence and in particular to determine the relative importance of:
i) idiosyncratic species-level change (random innovation) against clade-level reorientation of the directions of variation (specific innovation),
and ii) evolution along versus away from major axes of variation.
The contributions of these multivariate evolutionary pathways are poorly-known despite their importance for understanding the origins of diversity from macro- to mega-evolutionary scales. 
Here we assess the stability of trait correlations across the global radiation of birds and estimate and compare major axes of beak shape evolution across lineages and scales to address the contributions of elaboration and innovation to the generation of avian phenotypic diversity.

% paragraph 3: how we measure this (glmm) on the birds data + nestedness
\subsection{Modelling nested trait covariances}
Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires
1) large, multivariate datasets to allow exploration of trait covariances at different scales,
2) reliable and efficient computational methods to estimate the major axes of phenotypic variation,
%TG: Gavin has been going with "major axes of phenotypic divergence" throughout but I've changed it to "major axes of phenotypic variation". It's a semantic point and we can easily toggle between one or the other but I would argue that the LOTR captures more the main direction of (co)variation (i.e. what is the most likely path of changes) rather than divergence (that in my head reads more as varying away from something). 
and 3) a set of mathematical tools that can estimate degrees of elaboration and both random and specific innovation at any scale.
To meet the first challenges we use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
To meet the second two challenges we propose a novel analytical pipeline for measuring elaboration and random innovation at the species level and both elaboration and specific innovation at the clade level (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
In order to provide a baseline on which to calculate elaboration and innovation (i.e. the major axis of variation) we use phylogenetic generalised mixed effect models (pGLMM) with beak shape as an eight-dimensional response variable (see Methods; \cite{MCMCglmm}).
Such models allow us to measure variance-covariance matrices partitioned into phylogenetic and non-phylogenetic components. 
From these models we can calculate major axes of variation as the leading eigenvector of the variance-covariance matrix in trait space.
We regard the major axis of variation as a deep-time analogue of the microevolutionary \textbf{G} matrix and posit that it represents the line of least evolutionary resistance capturing the effects of historical contingency on multivariate evolution.
The major axis can be defined at any phylogenetic scale (i.e. for all species in the phylogeny, or for all species within any monophyletic clade).
In order to assess the roles of elaboration and innovation across scales we fit our pGLMM as nested models in which we define variance-covariance matrices (and therefore major axes of variation) for
i) the entire phylogeny of all species included in the species (hereafter the "whole tree" level),
ii) all species within each of nine clades mapping approximately to super-orders (the super-order level),
and iii) all species within each of 27 clades mapping to taxonomic orders (the order level).
These nested partitions of multivariate trait space provide the basis for all subsequent analyses.

\subsection{Elaboration dominates species-level divergence}
% paragraph 4: results for elaboration and innovation for species
We then measure innovation and elaboration as the multilinear algebraic projection of each species onto their nested major axis of variation.
Here are the results for each species. \ref{Fig:phylogeny}
There seems to be a nested structure of the variation in novelty (i.e. elaboration plus innovation).
There is more variation the deeper you baseline comparison in the tree is, suggesting that novelty is relative to the baseline comparison.
Also, the majority of that novelty is through elaboration.
We found that high levels of elaboration are more common across all birds than high levels of innovation (Fig. \ref{Fig:phylogeny}).
However, we also found various combinations of elaboration and innovation across birds.
For example, compared to the phylogenetic major axis, the Trochilidae (hummingbirds, with long narrow beaks) display high levels of elaboration and low innovation whereas the Bucerotidae (hornbills, with large and oddly shaped beaks) display high levels of innovation and low elaboration.
Other clades such as the Apodidae (swifts, with large but gaping beaks) and the distantly related but convergent Hirundidae (swallows and martins) display high levels of both elaboration and innovation.
Overall, we find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp?
The variability of, and apparent disconnect between, innovation and elaboration holds whether comparisons are made at the order, superorder, or class-wide phylogenetic level (Fig. \ref{Fig:phylogeny})%{#fig_correlations}),%TODO: maybe add to supp?
 and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
The dominance of elaboration when measured at the species level is consistent with ideas of evolution along lines of least evolutionary resistance, yet contrasts with our observations of frequent innovation via reorientation of trait space within clades.
\textbf{But there is a paradox here:}
If most organisms evolve incrementally along major axis of variation then how can we explain the staggering diversity of species morphologies across the tree of life?

@@

% Paragraph 5+: Some ellipses stuff with bird names.
However, this change in the line of least resistance's orientation between different clade levels varies between clades (Fig.\ref{Fig:ellipses}).
While orthogonality provides a simple description of the degree of reorientation of phenotypic axes (measuring the two-dimensional plane of an angle), it does not fully capture the multi-dimensionality of the data.
To further quantify the tension implied between theory on lines of least evolutionary resistance and observations of deep-time reorientation of trait covariance, we measured the elaboration and innovation of each different clades (\textit{cf.} the elaboration and innovation of a species - see below).
As noted above, elaboration describes multivariate evolution that is primarily along the phylogenetically constrained phenotypic major axis whereas innovation describes multivariate evolution that is primarily away from the phylogenetically constrained phenotypic major axis \cite{endler2005animal}.
Clades (or species) can be both elaborators and innovators.
To measure innovation and elaboration we translated the major axes of each clade's (i.e. superorders and orders) phylogenetic variance-covariance matrices onto the global phylogenetic major axis so that they shared the same origin in the shapespace.
We then used multilinear algebra to project a focal clade's major axis onto the phylogenetic major axis (Fig.S\ref{Fig:cheat_sheet}).
We interpret the measured projection (distance onto the phylogenetic major axis) as the clade's elaboration score and the measured rejection (distance from the major axis) as the clade's innovation score.
These scores were measured for each of 4000 posterior pairs of clade vs. phylogenetic variance-covariance matrix (Fig.\ref{Fig:ellipses} boxplots).
We found that for most clades the phenotypic major axis is not aligned with the global phylogenetic major axis, (Fig. \ref{Fig:ellipses}).
Remarkably, more than half of the assessed clades (4/8 superorders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to theory \cite{schluter1996adaptive,marroig2005size}, innovation is a more common generator of morphological diversity than elaboration.
However, there is substantial variation in terms of elaboration and innovation across clades including relatively low elaboration and innovation (e.g. Cuculiformes; Fig. \ref{Fig:ellipses}), low elaboration and high innovation (e.g. Galliformes; Fig. \ref{Fig:ellipses}), high elaboration and low innovation (e.g. Coraciiformes; Fig. \ref{Fig:ellipses}) or both relatively high elaboration and innovation (e.g. Podicipediformes; Fig. \ref{Fig:ellipses}).
These patterns hold across scales including within superorders and at finer scales within the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes}).
This suggests that divergence away from line of least evolutionary resistance (i.e. the phylogenetic major axis) is a common feature of avian beak evolution across scales and further supports the suggestion of a remarkable flexibility in the direction of evolution in deep-time even in a developmentally constrained trait.
